cardinalR: Generating interesting high-dimensional data structures

### 2025-08-24 paper-cardinalR.tex
Updated cross reference in sub sections

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in \(p\text{-}D\). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in \(p\text{-}D\). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

@@

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\). Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_orgcurvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in a \(p\)-dimensional space. Each branch is constructed in a unique or repeated \(2\text{-}D\) subspace of the \(p\)-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining \(p - 2\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_gridcube(n,\ p)} constructs a regular grid using approximately \(n\) points by evenly partitioning each of the \(p\) dimensions. Specifically, the number of grid points along each axis is determined by \texttt{gen\_nproduct(n,\ p)}, which calculates the nearest balanced integer factors whose product is close to \(n\). The full grid is generated via \texttt{expand\_grid()}, giving Cartesian combinations of these evenly spaced positions.

@@

The \texttt{gen\_unifcube(n,\ p)} function generates \(n\) points uniformly distributed within a \(3\text{-}D\) cube centered at the origin, optionally embedded in \(p\text{-}D\) with additional noise dimensions.

@@

The \texttt{gen\_cubehole(n,\ p)} function generates a dataset of \(n\) points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in \(p\text{-}D\).

@@

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions. This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions. This function creates a more complex curvilinear structure than a simple parabola.

@@

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base. The pyramid is embedded in \(p\text{-}D\), with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l_x, l_y)\) to a smaller rectangular section with side lengths \((2r_t, 2r_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

@@

The \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a triangular cross-section. The structure is embedded in a \(p\text{-}D\), with the tip located at height \(0\) and the base at height \(h\). The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

@@

The \texttt{gen\_pyrstar(n,\ p,\ h,\ rb)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-shaped structure with a hexagonal base. The pyramid extends vertically from height \(z = 0\) (tip) to height \(z = h\) (base), embedded in a \(p\)-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

@@

The \texttt{gen\_pyrholes(n,\ p)} function generates \(n\) points embedded in a \(p\text{-}D\) simplex using a chaotic attractor-like midpoint algorithm. The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a ``Sierpinski-like pyramid'' in high dimensions.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \(n\) observations in \(p\text{-}D\).

@@

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)}. This function produces \(n\) observations in \(p\text{-}D\) where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

@@

The function \texttt{gen\_circle(n,\ p)} generates a \(p\)-dimensional dataset of \(n\) observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale.

@@

The \texttt{gen\_curvycycle(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations. The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

@@

The \texttt{gen\_unifsphere(n,\ p,\ r)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on the surface of a \(3\text{-}D\) sphere of radius \(r\), with additional Gaussian noise dimensions added when \(p > 3\).

@@

The \texttt{gen\_gridedsphere(n,\ p)} function generates a \(p\)-dimensional dataset of approximately \(n\) points evenly distributed on the surface of a \((p!-!1)\)-sphere (the unit sphere embedded in \(\mathbb{R}^p\)).

@@

The \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates a synthetic dataset of \(n_1 + kn_2\) observations in \(p\)-dimensional space, consisting of one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location.

@@

The \texttt{gen\_hemisphere(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on a \(4\text{-}D\) hemisphere, optionally extended with Gaussian noise in additional dimensions when \(p > 4\).

@@

The \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise.

@@

The \texttt{gen\_curvycylinder(n,\ p,\ h)} function generates a \(p\)-dimensional dataset of \(n\) observations structured as a \(3\text{-}D\) cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when \(p > 4\).

@@

The \texttt{gen\_sphericalspiral(n,\ p,\ spins)} function simulates a dataset of \(n\) observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond \(4\text{-}D\), structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds.

@@

The \texttt{gen\_helicalspiral(n,\ p)} function generates a dataset of \(n\) observations forming a helical spiral, embedded in the first four dimensions of \(p\text{-}D\). Additional dimensions (if \(p > 4\)) are filled with structured noise to mimic high-dimensional complexity.

@@

The \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of \(p\text{-}D\). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix.

@@

The \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables.

@@

The function \texttt{gen\_trefoil4d()} generates a \(4\text{-}D\) trefoil knot using two angular parameters, \(\theta\) and \(\phi\). The knot is constructed on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)), with a band of thickness around the knot path, controlled by \texttt{steps}.

@@

The function \texttt{gen\_trefoil3d()} generates a \(3\text{-}D\) stereographic projection of the \(4\text{-}D\) trefoil knot by mapping each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using the transformation \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), while excluding points where \(X_4 = 1\) to avoid division by zero.

>>>>

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in \(p\text{-}D\) (Figure \ref{fig:branch-proj} a). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

@@

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in \(p\text{-}D\) (Figure \ref{fig:branch-proj} b). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

@@

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} d). Each branch lies primarily within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_orgcurvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in a \(p\)-dimensional space (Figure \ref{fig:branch-proj} e). Each branch is constructed in a unique or repeated \(2\text{-}D\) subspace of the \(p\)-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining \(p - 2\) dimensions contain Gaussian noise.

@@

The \texttt{gen\_gridcube(n,\ p)} constructs a regular grid using approximately \(n\) points by evenly partitioning each of the \(p\) dimensions (Figure \ref{fig:cube-proj} a). Specifically, the number of grid points along each axis is determined by \texttt{gen\_nproduct(n,\ p)}, which calculates the nearest balanced integer factors whose product is close to \(n\). The full grid is generated via \texttt{expand\_grid()}, giving Cartesian combinations of these evenly spaced positions.

@@

The \texttt{gen\_unifcube(n,\ p)} function generates \(n\) points uniformly distributed within a \(3\text{-}D\) cube centered at the origin, optionally embedded in \(p\text{-}D\) with additional noise dimensions (Figure \ref{fig:cube-proj} b).

@@

The \texttt{gen\_cubehole(n,\ p)} function generates a dataset of \(n\) points sampled uniformly inside a cube, then removes points that fall within a central spherical hole, resulting in a hollow cube structure in \(p\text{-}D\) (Figure \ref{fig:cube-proj} c).

@@

The \texttt{gen\_quadratic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a quadratic curve in the first two dimensions (Figure \ref{fig:poly-proj} a). This \(2\text{-}D\) parabolic structure is embedded within a higher-dimensional space with additive noise in the remaining dimensions.

@@

The \texttt{gen\_cubic(n,\ p,\ range)} function generates a dataset of \(n\) points forming a cubic curve in the first two dimensions (Figure \ref{fig:poly-proj} b). This function creates a more complex curvilinear structure than a simple parabola.

@@

The \texttt{gen\_pyrrect(n,\ p,\ h,\ l\_vec,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a rectangular base (Figure \ref{fig:pyr-proj} a). The pyramid is embedded in \(p\text{-}D\), with a tip at height zero and base at height \(h\). The shape tapers linearly from the base dimensions \((l_x, l_y)\) to a smaller rectangular section with side lengths \((2r_t, 2r_t)\) at the tip. Points are distributed more densely near the tip, simulating a natural skew toward smaller height values.

@@

The \texttt{gen\_pyrtri(n,\ p,\ h,\ l,\ rt)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-like structure with a triangular cross-section (Figure \ref{fig:pyr-proj} b). The structure is embedded in a \(p\text{-}D\), with the tip located at height \(0\) and the base at height \(h\). The triangle expands linearly in size from tip to base, with more points concentrated near the tip.

@@

The \texttt{gen\_pyrstar(n,\ p,\ h,\ rb)} function generates a dataset of \(n\) points forming a high-dimensional pyramid-shaped structure with a hexagonal base (Figure \ref{fig:pyr-proj} c). The pyramid extends vertically from height \(z = 0\) (tip) to height \(z = h\) (base), embedded in a \(p\)-dimensional space. The distribution concentrates more points near the base, and the hexagonal spread tapers toward the tip.

@@

The \texttt{gen\_pyrholes(n,\ p)} function generates \(n\) points embedded in a \(p\text{-}D\) simplex using a chaotic attractor-like midpoint algorithm (Figure \ref{fig:pyr-proj} d). The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a ``Sierpinski-like pyramid'' in high dimensions.

@@

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \(n\) observations in \(p\text{-}D\) (Figure \ref{fig:scurve-proj} a).

@@

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)} (Figure \ref{fig:scurve-proj} b). This function produces \(n\) observations in \(p\text{-}D\) where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

@@

The function \texttt{gen\_circle(n,\ p)} generates a \(p\)-dimensional dataset of \(n\) observations, where the first two dimensions form a unit circle, and the remaining dimensions are structured sinusoidal extensions of the angular parameter with progressively smaller scale (Figure \ref{fig:sphere-proj} a).

@@

The \texttt{gen\_curvycycle(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations lying on a curved closed loop with controlled high-dimensional sinusoidal deviations (Figure \ref{fig:sphere-proj} b). The structure forms a one-dimensional nonlinear cycle embedded in higher-dimensional space.

@@

The \texttt{gen\_unifsphere(n,\ p,\ r)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on the surface of a \(3\text{-}D\) sphere of radius \(r\), with additional Gaussian noise dimensions added when \(p > 3\) (Figure \ref{fig:sphere-proj} c).

@@

The \texttt{gen\_gridedsphere(n,\ p)} function generates a \(p\)-dimensional dataset of approximately \(n\) points evenly distributed on the surface of a \((p!-!1)\)-sphere (the unit sphere embedded in \(\mathbb{R}^p\)) (Figure \ref{fig:sphere-proj} d).

@@

The \texttt{gen\_clusteredspheres(n,\ k,\ p,\ r,\ loc)} function generates a synthetic dataset of \(n_1 + kn_2\) observations in \(p\)-dimensional space, consisting of one large sphere of radius \(r_1\) and \(k\) smaller spheres of radius \(r_2\), each centered at a different random location (Figure \ref{fig:sphere-proj} e).

@@

The \texttt{gen\_hemisphere(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations distributed approximately uniformly on a \(4\text{-}D\) hemisphere, optionally extended with Gaussian noise in additional dimensions when \(p > 4\) (Figure \ref{fig:sphere-proj} f).

@@

The \texttt{gen\_crescent(n,\ p)} function generates a \(p\)-dimensional dataset of \(n\) observations based on a \(2\text{-}D\) crescent-shaped manifold with optional structured high-dimensional noise (Figure \ref{fig:triginometric-proj} a).

@@

The \texttt{gen\_curvycylinder(n,\ p,\ h)} function generates a \(p\)-dimensional dataset of \(n\) observations structured as a \(3\text{-}D\) cylindrical manifold with an added nonlinear curvy dimension, and optional noise dimensions when \(p > 4\) (Figure \ref{fig:triginometric-proj} b).

@@

The \texttt{gen\_sphericalspiral(n,\ p,\ spins)} function simulates a dataset of \(n\) observations that form a spiraling path on a spherical surface in the first four dimensions. When extended beyond \(4\text{-}D\), structured nonlinear noise dimensions are added to simulate more realistic high-dimensional manifolds (Figure \ref{fig:triginometric-proj} c).

@@

The \texttt{gen\_helicalspiral(n,\ p)} function generates a dataset of \(n\) observations forming a helical spiral, embedded in the first four dimensions of \(p\text{-}D\). Additional dimensions (if \(p > 4\)) are filled with structured noise to mimic high-dimensional complexity (Figure \ref{fig:triginometric-proj} d).

@@

The \texttt{gen\_conicspiral(n,\ p,\ spins)} function generates a dataset of \(n\) points forming a conical spiral in the first four dimensions of \(p\text{-}D\) (Figure \ref{fig:triginometric-proj} e). The geometry combines radial expansion, vertical elevation, and spiral deformation, simulating a structure that fans out like a \(3\text{-}D\) conic helix.

@@

The \texttt{gen\_nonlinear(n,\ p,\ hc,\ non\_fac)} function simulates a non-linear \(2\text{-}D\) surface embedded in higher dimensions, constructed using inverse and trigonometric transformations applied to independent variables (Figure \ref{fig:triginometric-proj} f).

@@

The function \texttt{gen\_trefoil4d()} generates a \(4\text{-}D\) trefoil knot using two angular parameters, \(\theta\) and \(\phi\). The knot is constructed on the \(3\)-sphere (\(S^3 \subset \mathbb{R}^4\)), with a band of thickness around the knot path, controlled by \texttt{steps} (Figure \ref{fig:trefoil-proj} a).

@@

The function \texttt{gen\_trefoil3d()} generates a \(3\text{-}D\) stereographic projection of the \(4\text{-}D\) trefoil knot by mapping each point \((X_1, X_2, X_3, X_4) \in \mathbb{R}^4\) to \((X_1', X_2', X_3') \in \mathbb{R}^3\) using the transformation \(X_1' = X_1 / (1 - X_4)\), \(X_2' = X_2 / (1 - X_4)\), and \(X_3' = X_3 / (1 - X_4)\), while excluding points where \(X_4 = 1\) to avoid division by zero (Figure \ref{fig:trefoil-proj} b).
